Supplementary material for: non-coding RNAs of bird genomes

### 2013-11-11 paper/manuscript.tex
Updated figures and supp. mat.

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  Although we show a
significant number of lncRNAs are surprisingly well conserved between
birds and mammals including several intriguing cases where the
reported mammalian lncRNA function is not conserved in birds.  In
contrast, we show extensive conservation of classical ncRNAs (e.g.,
tRNAs) and more recently discovered ncRNAs (e.g., snoRNAs
and miRNAs) in birds, we also demonstrate apparent ``losses'' in
several RNA families. These include the divergence of some classical
ncRNAs and the loss of several snoRNAs and microRNAs.  These combined
results illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. We focus our report on the unusual
results within the avian lineages. These are either unexpectedly
well-conserved RNAs or unexpectedly poorly-conserved RNAs. The former
are RNAs we would not have expected to be conserved between the birds
and the organisms these genes were initially identified in; Usually,
this hypothesis is based upon the function of the RNA which is not
conserved in avian species. The latter are apparent losses of RNA
genes that were expected to be conserved; Usually, this hypothesis is
based upon the conservation of these RNAs in other vertebrate species.
We use three models of ``loss'' that explain the data: Firstly, these
could be genuine cases of an ancestral gene-loss along the avian
lineage. Secondly, this could be a case of ``divergence'' where a RNA
gene has undergone significant sequence and structural alterations, so
much so, that our homology search tools can no longer detect a
relationship between vertebrate exemplars and the avian
varieties. Thirdly, we consider the possibility that the available
genome assemblies have independently failed to capture these genes.

@@

In addition, the lncRNA DLEU2, is well conserved conserved across the
vertebrates, it is a host gene for two miRNA genes, miR-15 and miR-16,
both of which are also well conserved across the vertebrates (See
Supplemental Figure~2). DLEU2 is thought to be a tumor-suppressor gene
as it is frequently deleted in malignant tumours
\cite{Lerner:2009,Klein:2010}.

>>>>

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds.  We also
demonstrate extensive conservation of classical ncRNAs (e.g., tRNAs)
and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
birds. Furthermore, we have discovered apparent ``losses'' in several
RNA families, these include the divergence of some classical ncRNAs
and the loss of several snoRNAs and microRNAs.  These combined results
illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. We focus our report on the unusual
results within the avian lineages. These are either unexpectedly
well-conserved RNAs or unexpectedly poorly-conserved RNAs. The former
are RNAs we would not have expected to be conserved between the birds
and the organisms these genes were initially identified in; Usually,
this hypothesis is based upon the function of the RNA which is not
conserved in avian species. The latter are apparent losses of RNA
genes that were expected to be conserved; usually, this hypothesis is
based upon the conservation of these RNAs in other vertebrate species.
We use three models of ``loss'' that explain the data: Firstly, these
could be genuine cases of an ancestral gene-loss along the avian
lineage. Secondly, this could be a case of ``divergence'' where a RNA
gene has undergone significant sequence and structural alterations, so
much so, that our homology search tools can no longer detect a
relationship between vertebrate exemplars and the avian
varieties. Thirdly, we consider the possibility that the available
genome assemblies have independently failed to capture these genes.

@@

In addition, the lncRNA DLEU2 is well conserved across the
vertebrates, it is a host gene for two miRNA genes, miR-15 and miR-16,
both of which are also well conserved across the vertebrates (See
Supplemental Figure~2). DLEU2 is thought to be a tumor-suppressor gene
as it is frequently deleted in malignant tumours
\cite{Lerner:2009,Klein:2010}.

### 2013-11-11 paper/supplementary-material.tex
Updated figures and supp. mat.

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/Y.pdf}
  \caption[]{...}\label{fig:6}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/mir-17-cluster.pdf}
  \caption[]{...}\label{fig:8}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.85\textwidth]{figures/let-7-cluster.pdf}
  \caption[]{...}\label{fig:9}
\end{figure}

>>>>

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/Y.pdf}
  \caption[]{...}\label{fig:6}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/mir-17-cluster.pdf}
  \caption[Presence/absence table of mir-17 miRNA clusters]{This
    figure illustrates the distribution of all members of the mir-17
    clusters across birds, compared to fish and mammals. Columns
    correspond to the single miRNA sequences, ordered by cluster and
    position within its cluster. Rows correspond to the species and
    the cells contain the number of copies of the miRNA in the
    respective species. Colors correspond to orthologous miRNA
    families: mir-17 (orange), mir-19 (blue) and mir-92/25
    (green).There are 3 clusters, I-1 and I-X, and cluster II.  While
    the two copies of cluster I are completely conserved, cluster II
    has been lost as a whole in birds.}\label{fig:8}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.95\textwidth]{figures/let-7-cluster.pdf}
  \caption[Presence/absence table of let-7 miRNA clusters]{This figure
    illustrates the distribution of the let-7 paralogs and clustered
    miRNAs of the families mir-100 and mir-125. Columns correspond to
    the single miRNA sequences, ordered by cluster and position within
    its cluster. Rows correspond to the species and the cells contain
    the number of copies of the miRNA in the respective
    species. Cluster A, which is strongly conserved in vertebrates has
    been completely lost in the avian lineage. Another obvious loss in
    birds is cluster F. However, in tinamou we detected an unclustered
    miRNA that best fits to F-let-7-1 in sequence and structure, while
    we could not find an I-let-7 miRNA sequence in this
    species.}\label{fig:9}
\end{figure}
